Income and Food Price Transmission

### 2016-08-11 figures.Rnw
Initial changes in connection with AE rejection
JAE sub. vers.

\SweaveInput{data_prep.Rnw}

\section{Figures}
<<pt, echo=FALSE, include=FALSE, fig=TRUE, height=6, width=8>>=
a = 1.3
b = 1.2
par(mfrow=c(1,2))
curve(1-a^(-x), 0, 10, lty=1, xlab = "Income", ylab = "", ylim = c(0,1))
curve(b^(-x), add = T, col = "blue", lty=2)
title(main = "Shares", cex.main = 1)
legend("topright", c(expression(beta(y)), expression(gamma(y))),lty = c(2,1), col = c("blue", "black"))
curve((1-a^(-x))*b^(-x), 0, 10, xlab = "Income", ylab = "")
title(main = "Price transmission", cex.main = 1)
legend("topright", c(expression(theta(y))), lty = 1, col = "black")
@

<<gamma, echo=FALSE, include=FALSE, fig=TRUE, height=6, width=8>>=
library(ggplot2)
library("foreign")

dat = read.csv("dist_trade.csv", sep = ";")
dat$gamma = (dat$vop_covx+dat$vop_covm)/dat$vop_covt*100
ggplot(dat, aes(log(gdppcp00), gamma)) + 
  geom_point() + 
  geom_smooth() +
  theme_bw() +
  xlab("Log of GDP") +
  ylab(expression(paste("Estimated ", gamma))) +
  theme(panel.grid.major = element_line(colour = "grey"))
@

<<infl, echo=FALSE, include=FALSE, fig=TRUE, height=6, width=8>>=
library(gridExtra)

p1 = ggplot(aes(y = value, x = factor(time, labels = c("2007:1-2008:6", "2010:1-2010:12")), 
                fill = factor(variable, labels = c("FPI inflation", "CPI inflation", "Real FPI growth"))), data = infl_long1) + 
  geom_boxplot(outlier.shape = 1) +
  theme_bw() +
  scale_fill_grey(start = 0, end = 1) +
  theme(legend.position=c(.75,.775)) +
  labs(fill = "") +
  ylab("Annualized growth rates (%)") + 
  xlab("")

p2 = ggplot(aes(x=foodinfl78, y=foodinfl1011), data = infl) + 
  geom_point(shape=1) +
  geom_smooth(method=loess) +
  theme_bw() +
  ylab("FPI inflation 2010-11") + 
  xlab("FPI inflation 2007-08") +
  geom_text(aes(label=ifelse(foodinfl78>40,as.character(iso3c),'')),hjust=1.2,just=-1.2, 
            size=3) +
  geom_text(aes(label=ifelse(foodinfl1011>20,as.character(iso3c),'')),hjust=1.2,just=-1.2,
            size=3)

p3 = ggplot(aes(x=cpiinfl78, y=cpiinfl1011), data = infl) +
  geom_point(shape=1) +
  geom_smooth(method=loess) +
  theme_bw() +
  ylab("CPI inflation 2010-11") + 
  xlab("CPI inflation 2007-08") +
  geom_text(aes(label=ifelse(cpiinfl78>30,as.character(iso3c),'')),hjust=1.2,just=-1.2, 
            size=3) +
  geom_text(aes(label=ifelse(cpiinfl1011>15,as.character(iso3c),'')),hjust=1.2,just=-1.2,
            size=3)

p4 = ggplot(aes(x=rfpigr78, y=rfpigr1011), data = infl) + 
  geom_point(shape=1) +
  geom_smooth(method=loess) +
  theme_bw() +
  ylab("Real FPI growth 2010-11") + 
  xlab("Real FPI Growth 2007-08") +
  geom_text(aes(label=ifelse(rfpigr78>12,as.character(iso3c),'')),hjust=1.2,just=-1.2, 
            size=3) +
  geom_text(aes(label=ifelse(rfpigr1011>10,as.character(iso3c),'')),hjust=1.2,just=-1.2,
            size=3)

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width=1\textwidth]{Master-pt.pdf}
\caption{The effect of income on price transmission}
\label{fig:pt}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=1\textwidth]{Master-gamma.pdf}
\caption{Estimates of $\gamma$. Source: \cite{Anderson2008}. Own calculations.}
\label{fig:gamma}
\end{center}
\end{figure}

>>>>

\section{Figures}
<<infl, echo=FALSE, include=FALSE, fig=TRUE, height=6, width=8>>=
library(gridExtra)
library(ggplot2)

load(paste0(str, "infl_l1.Rda"))
load(paste0(str, "infl_0708_1011.Rda"))

p1 = ggplot(aes(y = value, x = factor(period, labels = c("2007:1-2008:6", "2010:1-2010:12")),
                fill = factor(variable, labels = c("FPI inflation", "CPI inflation", "Real FPI growth"))), data = infl.l1) + 
  geom_boxplot(outlier.shape = 1) +
  theme_bw() +
  scale_fill_grey(start = 0, end = 1) +
  theme(legend.position=c(.75,.775)) +
  labs(fill = "") +
  ylab("Annualized growth rates (%)") + 
  xlab("")

p2 = ggplot(aes(x=fpigr.0708, y=fpigr.1011), data = infl) + 
  geom_point(shape=1) +
  geom_smooth(method=loess) +
  theme_bw() +
  ylab("FPI inflation 2010-11") + 
  xlab("FPI inflation 2007-08") +
  geom_text(aes(label=ifelse(fpigr.0708>40,as.character(iso3c),'')),hjust=1.2,vjust=-1.2, 
            size=3) +
  geom_text(aes(label=ifelse(fpigr.1011>20,as.character(iso3c),'')),hjust=1.2,vjust=-1.2,
            size=3)

p3 = ggplot(aes(x=cpigr.0708, y=cpigr.1011), data = infl) +
  geom_point(shape=1) +
  geom_smooth(method=loess) +
  theme_bw() +
  ylab("CPI inflation 2010-11") + 
  xlab("CPI inflation 2007-08") +
  geom_text(aes(label=ifelse(cpigr.0708>30,as.character(iso3c),'')),hjust=1.2,vjust=-1.2, 
            size=3) +
  geom_text(aes(label=ifelse(cpigr.1011>15,as.character(iso3c),'')),hjust=1.2,vjust=-1.2,
            size=3)

p4 = ggplot(aes(x=rfpigr.0708, y=rfpigr.1011), data = infl) + 
  geom_point(shape=1) +
  geom_smooth(method=loess) +
  theme_bw() +
  ylab("Real FPI growth 2010-11") + 
  xlab("Real FPI Growth 2007-08") +
  geom_text(aes(label=ifelse(rfpigr.0708>12,as.character(iso3c),'')),hjust=1.2,vjust=-1.2, 
            size=3) +
  geom_text(aes(label=ifelse(rfpigr.1011>10,as.character(iso3c),'')),hjust=1.2,vjust=-1.2,
            size=3)

@@

### 2016-08-11 intro.Rnw
JAE sub. vers.

% !Rnw root = master.Rnw

\section{Introduction}
In response to the substantial increase in international food prices during the first decade of the new millennium, a large number of recent studies have quantified global and local poverty impacts of international price shocks.\footnote{E.g. \cite{Attanasio2013,Balagtas2014,Coxhead2012,Ferreira2013,Hoyos2011,Ivanic2012,Janvry2010,Mghenyi2011,Vu2011,Wodon2010}. In addition, a supplement to volume 39 of \emph{Agricultural Economics} is entirely devoted to this topic \citep[see][]{masters08}.} A major determinant of a country's vulnerability to such shocks is the responsiveness of its domestic food prices \citep[e.g.][]{Hoyos2011}. But what are the main determinants of the domestic price responsiveness to international shocks? Are they related to trade, consumption, geography or something else? As the title of the paper suggests, a country's income level seems to be an important determinant but other factors also play a role. In particular, countries which control their international trade through an agricultural state trading enterprise are found to experience lower food price transmission on average in the short run as well as in the long run.

The literature on domestic price impacts of international commodity price shocks is vast and the topic has experienced a revival in the wake of the 2007-08 food crisis \citep[see][]{Fackler2001, kouyate16}. However, since the majority of price transmission studies focus on a single country we know little about the determinants, other than distance and borders, of price transmission differences across countries. Moreover, most price transmission studies focus on individual commodity prices rather than food prices in general. This paper therefore contributes with a cross country analysis of price transmission determinants that is based on aggregate food price data, i.e. food price indexes, rather than prices of individual agricultural commodities.

Regarding the choice between individual or aggregate prices, each approach has its strengths and weaknesses \citep{headey10}. Commodity specific analyses have a stronger microeconomic foundation in the law of one price but results do not generalize to other food goods or even to the same food good traded in other domestic markets. Transmission estimates based on aggregate price data, on the other hand, give a more comprehensive picture and therefore better reflect domestic welfare costs of international food price fluctuations. A potential weakness though of the latter approach, is that a country's food price index might not reflect the composition of poor people's food baskets which is often based mainly on grain. Another weakness is that the composition of a country's food price index differs substantially from that of the international food price indexes and as well as from other countries' price indexes, which makes it more difficult to compare transmission rates across countries. These idiosyncrasies do not, however, preclude us from studying broad patterns in price transmission rates.

The empirical analysis is based on insights derived from a simple economic model that is more suitable for aggregate price data than the law of one price workhorse model. A key prediction is that a country's income level should have an independent effect on price transmission, in addition to factors related to trade openness, geography and so on. The model predicts a novel inverse-U shaped relationship between a country's income level and its price transmission rate. This income-price transmission relationship turns out to be a robust feature of the data.

>>>>

% !Rnw root = Master.Rnw

\section{Introduction}
In response to the substantial increase in international food prices during the first decade of the new millennium, a large number of recent studies have quantified global and local poverty impacts of international price shocks.\footnote{E.g. \cite{Attanasio2013,Balagtas2014,Coxhead2012,Ferreira2013,Hatzenbuehler2016, Hoyos2011,Ivanic2012,Janvry2010,Masters2008, Mghenyi2011,Vu2011,Wodon2010}.} A major determinant of a country's vulnerability to such shocks is the responsiveness of its domestic food prices \citep[e.g.][]{Hoyos2011}. But what are the main determinants of the domestic price responsiveness to international shocks? Are they related to trade, consumption, geography or something else? As the title of the paper suggests, a country's income level is an important determinant but other factors also seem to play a role.

The literature on domestic price impacts of international commodity price shocks is vast and the topic has experienced a revival in the wake of the 2007-08 food crisis \citep[see][]{Fackler2001, Kouyate2016}. However, we know little about the determinants, other than distance and borders, of price transmission differences across countries. Moreover, most price transmission studies focus on individual commodity prices rather than food prices in general. This paper therefore contributes with a cross country analysis of price transmission determinants that is based on aggregate food price data, i.e. food price indexes, rather than prices of individual agricultural commodities.

Regarding the choice between individual or aggregate prices, each approach has its strengths and weaknesses \citep{Headey2010}. Commodity specific analyses have a stronger microeconomic foundation in the ``law of one price'' but results do not generalize to other food goods or even to the same food good traded in other domestic markets. Transmission estimates based on aggregate price data, on the other hand, give a more comprehensive picture and therefore better reflect domestic welfare costs of international food price fluctuations. A potential weakness though of the latter approach, is that a country's food price index might not reflect the composition of poor people's food baskets. Another weakness is that the composition of a country's food price index differs substantially from that of the international food price indexes and as well as from other countries' price indexes, which makes it more difficult to compare transmission rates across countries. These idiosyncrasies do not, however, preclude us from studying broad patterns in price transmission rates.

The empirical analysis is based on insights derived from a simple economic model that is more suitable for aggregate price data than the ``law of one price'' workhorse model. A key prediction is that a country's income level should have an independent effect on price transmission, in addition to factors related to trade openness, geography and so on. The model predicts a novel inverse-U shaped relationship between a country's income level and its price transmission rate. This income-price transmission relationship turns out to be a robust feature of the data.

### 2016-08-11 model.Rnw
JAE sub. vers.

Two studies quantify the determinants of price transmission and market integration based on individual agricultural commodity prices rather than food price indexes. \citet{kouyate16} carry out a meta-regression analysis of the effect of distance on agricultural commodity price transmission for a large number of market pairs. Their results show that geographical distance and separation by borders reduces the likelihood of price cointegration and weakens the speed of speed of price transmission significantly. These findings are consistent with the well established stylized facts from the trade literature, that borders and distance reduces trade.

Finally, \citet{greb12} consider 8 variables covering geographical, infrastructural, institutional and commodity specific factors which are believed to influence agricultural price transmission. The authors regress their own price transmission estimates, as well as estimates obtained from the literature, on these country characteristics. What they find is that the covariates explain more of the variation in the estimated speed of adjustment parameters (the $\alpha$'s) than in the cointegration parameters (the $\beta$'s). Another major finding is that more of these covariates have significant and plausible partial effects on the $\alpha$'s than on the $\beta$'s. However, none of the determinants have a robust robust impact on price transmission.

>>>>

Two studies quantify the determinants of price transmission and market integration based on individual agricultural commodity prices rather than food price indexes. \citet{Kouyate2016} carry out a meta-regression analysis of the effect of distance on agricultural commodity price transmission for a large number of market pairs. Their results show that geographical distance and separation by borders reduces the likelihood of price cointegration and weakens the speed of speed of price transmission significantly. These findings are consistent with the well established stylized facts from the trade literature, that borders and distance reduces trade.

Finally, \citet{Greb2016} consider 8 variables covering geographical, infrastructural, institutional and commodity specific factors which are believed to influence agricultural price transmission. The authors regress their own price transmission estimates, as well as estimates obtained from the literature, on these country characteristics. What they find is that the covariates explain more of the variation in the estimated speed of adjustment parameters (the $\alpha$'s) than in the cointegration parameters (the $\beta$'s). Another major finding is that more of these covariates have significant and plausible partial effects on the $\alpha$'s than on the $\beta$'s. However, none of the determinants have a robust robust impact on price transmission.
